|

### 2021-09-15 paper.Rmd
some progress

---
title: |
  | Emotional framing and the effectiveness of 
  | corrective information^[I thank Thomas Bräuninger, Sebastian Juhl, the GIP team, and participants of the Fall 2020 CDSS workshop in political science at the University of Mannheim for excellent comments and helpful suggestions on this study. This work was supported by the German Research Foundation (DFG) via the Collaborative Research Center (SFB) 884 on ``The Political Economy of Reforms'' (Project C2) and the University of Mannheim’s Graduate School of Economics and Social Sciences.]
  | *Working Paper*
author: |
  | *Working paper*
  |
  | Pirmin Stöckle^[Corresponding address: pirmin.stoeckle@gess.uni-mannheim.de]
  | GESS and SFB 884, University of Mannheim
date: \today
linestretch: 1.5
colorlinks: true
abstract: \setstretch{1} \noindent Concerns about various forms of misinformation and its fast dissemination through online media have generated huge interest into ways to effectively correct false claims. An under-explored mechanism in this research is the role of distinct emotions. How do emotional appeals interact with corrective information? Specifically, I focus on the emotion of disgust, which has been shown to be linked to the moralization of attitudes, which in turn reduces the impact of empirical evidence on attitudes and makes compromise less likely. Substantively, I investigate the issue of genetically modified (GM) food. I implement a pre-registered survey experiment within a panel study based on a probability sample of the general population in Germany (N = 4,000). Results show what we have learned about the question. This study provides evidence on the effect of emotionally charged disinformation on perceptions of GM food, and ways to effectively correct false claims. In a broader perspective, these results inform further studies and policy interventions on other issues where disinformation loads on strong emotions, ranging from social policy over immigration to health interventions such as vaccinations. \vspace{.8cm}
# abstract-template: Everyone agrees that this issue is really important. But we do not know much about this specific question, although it matters a great deal, for these reasons. We approach the problem from this perspective. Our research design focuses on these cases and relies on these data, which we analyze using this method. Results show what we have learned about the question. They have these broader implications.

@@

To investigate the direct effects of misinformation and corrective information conditional on emotional framing, I conducted a pre-registered survey experiment.^[Pre-registration can be found on OSF: https://osf.io/y63xu] The experiment was implemented within wave 53 of the German Internet Panel (GIP) [@blom2015setting]. The GIP is a longitudinal panel survey based on a probability sample of the general population in Germany. The survey is conducted online, but recruitment was conducted offline according to a strict random probability sample. Households that did not posses the technical devices to participate online have been equipped with such devices. Fieldwork for this study took place in May 2021 and the sample size reached 4,260 respondents.

The between-subjects survey experiment followed a 3 $\times$ 3 full factorial design varying both misinformation (none, low emotional framing, high emotional framing) and corrective information (none, neutral correction, counterframe correction). Respondents were randomly assigned to one of the 9 resulting groups, thus yielding 8 treatment conditions and one control condition were no information treatment was shown.

@@

```{r, warning=FALSE, echo=FALSE}
# load data
source("data-preparation.R")
```

@@

Table \@ref(tab:descriptives)

@@

To quantify the effects of the different treatment conditions on the outcomes of interest, I run separate linear regressions with the respective outcome as the dependent variable and the treatment group indicator as the independent variable.

@@

```{r fig-1, fig.align="center", fig.cap="Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
# H1: misinformation effect
pdata_h1_belief_policy <- rbind(get_pdata(lm_moral_conviction, c(4, 7), "Moral conviction"),
                                get_pdata(lm_policy_opinion, c(4, 7), "Policy restrictions"),
                                get_pdata(lm_knowledge_safe, c(4, 7), "Belief GMO safety"))

@@

pdata_h1 <- pdata_h1 %>% 
  mutate(emotion = ifelse(str_detect(group, "low-misinfo"), 
                          "Low emotional misinformation", "High emotional misinformation"))

# actual plotting
ggplot(pdata_h1, aes(x = outcome, y = estimate, color = "1")) +
  geom_point(size = 1.5, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = lower, ymax = upper, linetype = emotion),
                size = 0.5,
                width = 0.25,
                position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 0, linetype = 3) +
  ylim(c(-0.2, 0.2)) +
  labs(x = "",
       y = "") +
  facet_wrap(fct_rev(emotion) ~ fct_rev(type), scales = "free") +
  coord_flip() +
  #scale_colour_manual(values = c("grey25", "grey50", "grey75")) +
  scale_color_viridis_d(begin = 0, end = 0.5, direction = 1) +
  scale_linetype_manual(values=c("solid", "twodash")) +
  scale_shape_manual(values = c(17, 16)) +
  theme_nice() +   
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        legend.key.size = unit(1, 'cm'),
        strip.text.x = element_text(size = 8)
  ) +
  guides(color = "none",
         linetype = guide_legend(reverse = TRUE),
         shape = guide_legend(reverse = TRUE))

@@

pdata_neutral_correction <- pdata_neutral_correction %>% 
  mutate(emotion = ifelse(str_detect(group, "low-misinfo"), "Low emotional misinformation", "High emotional misinformation"),
         correction = ifelse(str_detect(group, "correction"), "Neutral correction", "No correction"))

@@

pdata_counter_frame <- pdata_counter_frame %>% 
  mutate(emotion = ifelse(str_detect(group, "low-misinfo"), "Low emotional misinformation", "High emotional misinformation"),
         correction = ifelse(str_detect(group, "counter"), "Counter-frame correction", "No correction"))

>>>>

---
title: |
  | Emotional framing and the effectiveness of 
  | corrective information^[I thank Thomas Bräuninger, Sebastian Juhl, the GIP team, and participants of the Fall 2020 CDSS workshop in political science at the University of Mannheim for excellent comments and helpful suggestions on this study. This work was supported by the German Research Foundation (DFG) via the Collaborative Research Center (SFB) 884 on ``The Political Economy of Reforms'' (Project C2) and the University of Mannheim’s Graduate School of Economics and Social Sciences.]
author: |
  | *Working paper*
  |
  | Pirmin Stöckle^[Corresponding address: pirmin.stoeckle@gess.uni-mannheim.de]
  | GESS and SFB 884, University of Mannheim
date: \today
linestretch: 1.5
colorlinks: true
abstract: \setstretch{1} \noindent Concerns about various forms of misinformation and its fast dissemination through online media have generated huge interest into ways to effectively correct false claims. An under-explored mechanism in this research is the role of distinct emotions. How do emotional appeals interact with corrective information? Specifically, I focus on the emotion of disgust, which has been shown to be linked to the moralization of attitudes, which in turn reduces the impact of empirical evidence on attitudes and makes compromise less likely. Substantively, I investigate the issue of genetically modified (GM) food. I implement a pre-registered survey experiment within a panel study based on a probability sample of the general population in Germany (N = 4,000). Results show what we have learned about the question. This study provides evidence on the effect of emotionally charged disinformation on perceptions of GM food, and ways to effectively correct false claims. In a broader perspective, these results inform further studies and policy interventions on other issues where disinformation loads on strong emotions, ranging from social policy over immigration to health interventions such as vaccinations. \vspace{.8cm}
# abstract-template: Everyone agrees that this issue is really important. But we do not know much about this specific question, although it matters a great deal, for these reasons. We approach the problem from this perspective. Our research design focuses on these cases and relies on these data, which we analyze using this method. Results show what we have learned about the question. They have these broader implications.

@@

To investigate the direct effects of misinformation and corrective information conditional on emotional framing, I conducted a pre-registered survey experiment.^[Pre-registration can be found on OSF: https://osf.io/y63xu.] The experiment was implemented within wave 53 of the German Internet Panel (GIP) [@blom2015setting]. The GIP is a longitudinal panel survey based on a probability sample of the general population in Germany. The survey is conducted online, but recruitment was conducted offline according to a strict random probability sample. Households that did not posses the technical devices to participate online have been equipped with such devices. Fieldwork for this study took place in May 2021 and the sample size reached 4,260 respondents.

The between-subjects survey experiment followed a 3 $\times$ 3 full factorial design varying both misinformation (none, low emotional framing, high emotional framing) and corrective information (none, neutral correction, counterframe correction). Respondents were randomly assigned to one of the nine resulting groups, thus yielding eight treatment conditions and one control condition were no information treatment was shown.

@@

```{r, warning=FALSE, echo=FALSE, message = FALSE}
# load data
source("data-preparation.R")
```

@@

Table \@ref(tab:descriptives) shows descriptive statistics of the outcome variables in the control group, which does not get any informational treatment and thus serves as the reference to investigate information effects. On the scale from 0 to 4, anger as an emotional reaction to th issue of GM food was expressed with a mean of 1.7 moderately below the midpoint. Anxiety was above the midpoint with a mean of 2.4 whereas disgust was realtively low with a mean of 1.3, and sadness again moderatly below the midpoint with 1.7. These figures indicate that respondents' negative reactions to GM food stem from concerns rather than disgust.

With regard to the factual belief that GM food is safe to eat, the sample is skewed toward the view that this is rather not true (mean = 3.2 on a scale from 1 to 7). When it comes to opinions about policy restrictions for GM food, respondents are rather in favor (mean = 5.2 on a scale from 1 to 7). Finally, the measure of moral conviction is close to the midpoint (mean = 4.4 on a scale from 0 to 8).

@@

I now turn to the effects of the informational treatments on the outcome variables. To quantify these effects, I run separate linear regressions with the respective outcome as the dependent variable and the treatment group indicator as the independent variable. For a more intuitive interpretation of the effect sizes - and to be able to compare outcome variables measured on different scales - I rescale all outcome variables to range from 0 to 1. I do not include any control variables because the random assignment of the experimental design provides the better safeguard against any possible confounders.

In the appendix, I report the regression tables, in the following I will visualize the estimated regression coefficients for the different treatment conditions.

@@

First, \@ref(fig:misinfo-only) reports the effects of reiceiving only the misinformation message. The upper half of the figure shows two plots with the effects of the low emotion misinformation and the lower half shows two plots with the effects of the high emtoion misinformation message. The y-axis indicates the different outomce variables, and the x-axis indicates the effect size relative to the control group, which serves as the reference category for all models.

```{r misinfo-only, fig.align="center", fig.cap="Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
# H1: misinformation effect
pdata_h1_belief_policy <- rbind(get_pdata(lm_moral_conviction, c(4, 7), "Moral conviction"),
                                get_pdata(lm_policy_opinion, c(4, 7), "Policy restrictions"),
                                get_pdata(lm_knowledge_safe, c(4, 7), "Belief GMO safety"))

@@

pdata_h1 <- pdata_h1 %>% 
  mutate(emotion = ifelse(str_detect(group, "low-misinfo"), 
                          "Low emotion misinformation", "High emotion misinformation"))

# actual plotting
ggplot(pdata_h1, aes(x = outcome, y = estimate, color = "1")) +
  geom_point(size = 1.5, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = lower, ymax = upper, linetype = emotion),
                size = 0.5,
                width = 0.25,
                position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 0, linetype = 3) +
  ylim(c(-0.2, 0.2)) +
  labs(x = "",
       y = "") +
  facet_wrap(fct_rev(emotion) ~ fct_rev(type), scales = "free") +
  coord_flip() +
  #scale_colour_manual(values = c("grey25", "grey50", "grey75")) +
  scale_color_viridis_d(begin = 0, end = 0.5, direction = 1) +
  scale_linetype_manual(values=c("solid", "twodash")) +
  scale_shape_manual(values = c(17, 16)) +
  theme_nice() +   
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        legend.key.size = unit(0.1, 'cm'),
        strip.text.x = element_text(size = 8)
  ) +
  guides(color = "none",
         linetype = guide_legend(reverse = TRUE),
         shape = guide_legend(reverse = TRUE))

@@

pdata_neutral_correction <- pdata_neutral_correction %>% 
  mutate(emotion = ifelse(str_detect(group, "low-misinfo"), "Low emotion misinformation", "High emotion misinformation"),
         correction = ifelse(str_detect(group, "correction"), "Neutral correction", "No correction"))

@@

pdata_counter_frame <- pdata_counter_frame %>% 
  mutate(emotion = ifelse(str_detect(group, "low-misinfo"), "Low emotion misinformation", "High emotion misinformation"),
         correction = ifelse(str_detect(group, "counter"), "Counterframe correction", "No correction"))
